---
title: "Serif Health Data Science Take Home"
author: 'Samantha Dean'
output: html_document
date: "2025-03-05"
---

```{r setup, include=FALSE, warning=F}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) #for data wrangling
```

## Workflow
- Spend some time looking into the data, making sure I understand the variables
- Check for and correct data QC issues that will cause issues with merging
- Exploratory analysis to understand how the two data sets fit together
- Construct merged dataset
- Write up

## Import Data
```{r import}
hpt = read.csv('rawdata/hpt_extract_20250213.csv')
tic = read.csv('rawdata/tic_extract_20250213.csv')
```

## Exploratory analysis and cleaning
I started with some exploratory analysis and recognized that making a data dictionary
at the start would save me time as I worked on this. I did some google-ing and made 
some quick tables of different columns to make sure I understood what all of the variables 
meant. The rough data dictionary is in the git repo. Through this process I identified 
some data abnormalities - namely that missing values were reading in as strings, 
payer names were formatted differently in each data set, and there were some incorrectly 
formatted codes. I made these fixes. When making fixes, I tried to make changes as general 
as possible, so that they could fix similar mistakes in future data deliveries.
```{r cleaning}
clean_hpt = hpt %>% 
  mutate(across(source_file_name:additional_generic_notes, ~ifelse(.=='null', NA, .)),                     # change nulls to NA
         across(standard_charge_gross:standard_charge_max, as.numeric),                                    # make rates numeric
         raw_code = as.integer(str_remove_all(hpt$raw_code, '[^0-9.]')),                                   # remove non numeric characters from codes
         clean_payer_name = str_to_lower(str_remove_all(str_remove_all(payer_name,"[ -]"), 'corporation')) # strip payer name formatting for easier matching
         )

clean_tic = tic %>%
  mutate(across(payer:cms_baseline_rate, ~ifelse(.=='null', NA, .)),
         across(c(rate, cms_baseline_rate), as.numeric),
         clean_payer_name = str_to_lower(str_remove_all(str_remove_all(payer,"[ -]"), 'corporation')))

```

## Formatting HPT data
I noticed that while all of the plans in the TIC dataset were open choice plans (ie 
included in and out of network providers), the HPT dataset included numerous other 
plan types. I wanted to make sure I compared rates with the same plan type across the 
two datasets, so I added in an indicator for the open choice plans in the HPT data. 
Then, I aggregated to the hospital / payer / code type / code / care setting / open 
indicator level. I observed that at this level of aggregation, even if there were 
multiple observations, rates were constant within HPT.

```{r hpt}
clean_hpt = clean_hpt %>% 
  mutate(open_plan = ifelse(str_detect(plan_name, 'open') | str_detect(str_to_lower(plan_name), 'all payer'), 1, 0))

# aggregate hpt to hospital / payer / code / setting level
agg_hpt = clean_hpt %>% 
  group_by(hospital_name, clean_payer_name, code_type, raw_code, setting, open_plan) %>% 
  summarise(hpt_count = n(), hpt_min = min(standard_charge_negotiated_dollar), hpt_max = max(standard_charge_negotiated_dollar))

```

## Formatting the TIC data
I extracted information from the cms_baseline_schedule variable to approximate the 
setting variable in the HPT dataset, by grouping into inpatient and outpatient. I 
excluded rows for non facility provider fees, because all of our hospital expenses 
explicitly are linked to a facility. This is an assumption I could check more closely with more time. 

Then I aggregated TIC data to the payer / code type / code / setting level. In this 
dataset, at this level there was a lot of variation still in observed prices. I produced 
summary statistics showing the min, q1, median, q3, and max rate values observed at this 
level. I am considering the TIC data to enumerate a distribution from which the HPT rate
can be considered a single drawn observation. 
```{r tic}
clean_tic = clean_tic %>%
  mutate(setting = case_when(cms_baseline_schedule == 'IPPS' ~ 'inpatient',
                             cms_baseline_schedule == 'OPPS' ~ 'outpatient',
                             str_detect(cms_baseline_schedule, 'PFS_FACILITY') ~ 'in facility provider fee',
                             str_detect(cms_baseline_schedule, 'PFS_NONFACILITY') ~ 'non facility provider fee')) %>%
  filter(setting != 'non facility provider fee') #because all the charges came from facilities in hpt

agg_tic = clean_tic %>% 
  group_by(clean_payer_name,code_type, code, setting) %>% 
  summarise(tic_count = n(), 
            tic_min = min(rate), 
            tic_q1 = quantile(rate, .25),
            tic_q2 = quantile(rate,.5),
            tic_q3 = quantile(rate,.75),
            tic_max = max(rate)) %>%
  ungroup(setting) %>%
  summarise(open_plan = 1, across(tic_count:tic_max, sum))
```

## Combine data sets
Now that I have both data sets aggregated to at least a payer / code type / code / setting 
level, I want to combine them. I produce two output products: one with all rows that include
pricing information from at least one of the source datasets, and one only with rows that 
include information from both TIC and HPT datasets. 
```{r merge}
# make empty data frame with the combinations of hospitals, codes, payer
id_cols = c('hospital_name', 'clean_payer_name', 'code_type', 'raw_code', 'open_plan')

full = expand.grid(unique(clean_hpt$hospital_name), 
                   unique(clean_hpt$clean_payer_name), 
                   unique(clean_hpt$code_type), 
                   unique(clean_hpt$raw_code),
                   0:1)
names(full) <- id_cols
  
full = full %>%
  merge(agg_hpt, by = id_cols, all = T) %>% 
  mutate(hpt_rate = hpt_min) %>%
  select(-hpt_min, -hpt_max) %>%
  merge(agg_tic, by.x = c(id_cols[-1]), by.y = c('clean_payer_name', 'code_type', 'code', 'open_plan'), all = T) %>%
  mutate(delta = hpt_rate - tic_q2,
         delta_pct = delta / hpt_rate)

# remove rows where there is no pricing information from either data set
full = full %>%
  filter(!(is.na(hpt_rate) & is.na(tic_q2)))
write.csv(full, 'outputdata/output_all.csv')

write.csv(full %>% filter(!(is.na(hpt_rate) | is.na(tic_q2))), 'outputdata/output_overlap.csv')

```

I identified 8 (out of a possible 36) unique hospital / code type / code / payer 
combinations that map closely between the TIC and the HPT dataset. At this level, 
the HPT dataset has a single, consistent rate repeated across multiple observations. 
The TIC data set has a larger number of observations, with a distribution of values 
at this level. I elected to present the single rate from the HPT dataset, and contextualize 
it with the min, max, and quartile values of the same aggregated level in the TIC data 
set. This approach can build trust through transparency and gives useful information on 
where a particular hospital's charge stands relative to the distribution of charges from 
other hospitals / claims. I also provided a delta column, calculating the difference 
between the consolidated HPT payment rate and the median rate from the corresponding 
TIC rates. The delta_pct column shows the percent change (/100) between the HPT and median 
TIC rates, and can be used as a measure of agreement between the two. 


## Assumptions and challenges
- I assumed that, within each payer, all the open access plans in HPT are the same 
thing, even with slightly different names. I checked that the rates were constant 
across the plan names, so this seems like a fine assunmption
- A tricky problem for me was how to reconcile the Setting variable (Inpatient / 
Outpatient / Both) from HIC with the CMS_Baseline_Schedule categories in the TIC 
data. I want to be sure I am mapping inpatient costs to inpatient costs and comparing
apples to apples, but not all observations were clearly categorized into inpatient / 
outpatient, especially for the TIC data. I made the assumption that provider fees 
were stacked on top of fees that were explicitly inpatient or outpatient in TIC, 
when both were within a single code type (eg CPT). 
- In the HPT notes fields, there was text that I could have scraped about the fee 
systems. I did not have time to work with the text fields, but specifically the 
"per diem" flag here could have been useful to consider for rates were not comparable 
to TIC rates. 
- I assumed that TIC data from open choice plans wouldn't be comparable to other 
kinds of plans in the HPT data. However, I would need to do some more research on 
insurance rates to check if this is an overly conservative assumption.
- I wasn't able to directly map the TIC data to individual hospitals. Maybe there 
is information available there where I could approximate which row of in TIC comes 
from which hospital, or even comes from a "similar" hospital. Since I couldnt be sure, 
I thought it made more sense to be straightforward that I am mapping a distribution 
from TIC to a single hospital in HPT. The ultimate goal, however, would be to make 
the "distribution" from TIC that I am comparing the HPT data to as specific as possible, 
so that I am only comparing within very similar strata.
- In this data extract, I didn't have to worry too much about run time. However, 
with the much larger full datasets, this could become a major consideration. Things 
like text field searching and other algorithmic processes can take a long time to 
run. With more time, I would optimize my approach for speed. 


